---
title: "Effects of Bear Endozoochory on Germination and Disperal"
author: "Clayton T. Lamb"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: github_document
---

```{r render, eval=FALSE,include=FALSE}
rmarkdown::render(here::here("huck_stats_glmm_only.Rmd"),
  output_file = "README.md"
)
```

# Purpose
Task 1: Analyze the effect of four experimental treatments on germination success of black huckleberry to better understand the effect of bear endozoochory. The four treatments were A) Whole Berry: a single, whole berry, B) Seeds from Berry: seeds extracted from a berry, C) Seeds from Scat: seeds extracted from scat, D) Mixed Scat: seeds and scat. Multiple seeds for each treatments were taken from the same sample (scat or sampling area), thus we included a random effect for sample ID to account for this nested data structure.

Task 2: Assess the movement rates of grizzly bears feeding in huckleberry patches to estimate potential seed dispersal.

## Load packages and data
```{r Load packages and data, results='hide', message=FALSE, warning=FALSE}
## load packages
library(renv)
library(readxl)
library(janitor)
library(lme4)
library(sjPlot)
library(ggeffects)
library(knitr)
library(hrbrthemes)
library(forcats)
library(sf)
library(lubridate)
library(terra)
library(emmeans)
library(multcomp)
library(tidyverse)


## Mixed effects modelling
df <- read_excel("data/Raw_Data_Huck_Thesis.xlsx", sheet = "60 Day Raw Data") %>%
  clean_names() %>%
  select(tray:total_seedlings) %>%
  mutate(days = 60) %>%
  rbind(
    read_excel("data/Raw_Data_Huck_Thesis.xlsx", sheet = "30 Day Raw Data") %>%
      clean_names() %>%
      select(tray:total_seedlings) %>%
      mutate(days = 30)
  ) %>%
  drop_na()
```
#Task 1: Analyze germination experiment
## Prep data for analysis
```{r prep,  message=FALSE, warning=FALSE}
## add in number of seeds used for each trial and trial names
seed.n <- tibble(
  treatment = c("A", "B", "C", "D"),
  n_seeds = c(27, 4, 4, 12),
  name= c("Whole berry", "Seeds from berry", "Seeds from scat", "Mixed scat")
)

df <- df %>%
  left_join(seed.n, by = "treatment") %>%
  mutate(no.germ = n_seeds - total_seedlings,
         name=fct_reorder(name, treatment%>%as.factor%>%as.numeric))%>%
  mutate(name_compress=str_remove_all(name," "))%>%
  mutate(name_compress=fct_relevel(name_compress,c("Wholeberry", "Seedsfromberry", "Seedsfromscat", "Mixedscat")))


## fix a few negatives
df <- df %>%
  mutate(total_seedlings = case_when(
    no.germ < 0 ~ n_seeds,
    TRUE ~ total_seedlings
  )) %>%
  mutate(
    no.germ = n_seeds - total_seedlings,
    prop.germ = total_seedlings / n_seeds
  )

##add days and treatment as a factor
df <- df%>%
  mutate(days.factor=paste0(as.factor(days), " days"))

##set up random effects
df <-df%>%
  mutate(jiffy=paste0(tray,number,letter),
         units=case_when(name%in%c("Seeds from berry", "Whole berry")~jiffy, TRUE~id))

## summary stats
summar.stats <- df %>%
  group_by(name, days.factor, treatment) %>%
  summarise(seedlings=sum(total_seedlings),
            seeds=sum(n_seeds),
            prop.germinated = (seedlings / seeds),
            perc.germinated = prop.germinated * 100,
            perc.germinated.bypot = mean(prop.germ) * 100,
            n=n(),
            se.prop=sqrt((prop.germinated*(1-prop.germinated))/n()),
            n_id=n_distinct(id),
            n_units=n_distinct(units),
            n_jiffy=n_distinct(jiffy)
            )

##plot
ggplot(summar.stats, aes(x=name, y=prop.germinated, ymin=prop.germinated-se.prop, ymax=prop.germinated+se.prop, fill=days.factor))+
  geom_col(position = position_dodge(width=0.9))+
  geom_linerange(position = position_dodge(width=0.9))+
  theme_ipsum()+
  labs(x="Treatment", y="Proportion germinated", fill="")+
  theme(legend.position = c(0.9,0.9),
        axis.title.x = element_text(size=17),
        axis.title.y = element_text(size=17),
        axis.text.x = element_text(size=13),
        axis.text.y = element_text(size=13),
        legend.text = element_text(size=13),
        plot.title = element_text(size=22),
        plot.subtitle = element_text(size=17))

ggsave("plots/barplot.png", width=8, heigh=5, units="in", dpi=300, bg="white")
```

## Model

```{r glm, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
### Start without random effect (i.e., not accounting for the nested structure of data [repeat samples from same berry or scat])
# fit model
m0 <- glm(prop.germ ~ name + days.factor, data = df, family = "binomial", weights = n_seeds)

# plot odds ratios
sjPlot::plot_model(m0,
  show.values = TRUE, show.p = TRUE,
  title = "Effect of Treatment on Germination"
)
# model table
sjPlot::tab_model(m0)

emmeans(m0, list(pairwise ~ name), adjust = "tukey")$`pairwise differences of name` %>% kable()
```

## with random effect accounting for the nested structure of data [repeat samples from same berry or scat)
```{r glmm,  message=FALSE, warning=FALSE}

# fit random effect model that accounts for id via a nested random effect of id on the intercept
m1 <- glmer(prop.germ ~ name_compress + days.factor + (1|id/jiffy), family = "binomial", weights = n_seeds, data = df)
summary(m1)

# plot odds ratios
sjPlot::plot_model(m1,
  show.values = TRUE, show.p = TRUE,
  title = "Effect of Treatment on Germination, w/ Random Effect for Sample"
)

# model table
sjPlot::tab_model(m1)

# pairwise significance
#emmeans(m1, list(pairwise ~ name_compress), adjust = "tukey")$`pairwise differences of name` %>% kable()
summary(glht(m1, linfct=c("name_compressSeedsfromberry - name_compressSeedsfromscat == 0",
                          "name_compressSeedsfromscat - name_compressMixedscat == 0")))

```

## test for days since
```{r glmm.dayssince,  message=FALSE, warning=FALSE}

# fit random effect model that accounts for id via a nested random effect of id on the intercept
m1_30 <- glmer(prop.germ ~ name_compress + (1|id/jiffy), family = "binomial", weights = n_seeds, data = df%>%filter(days==30))
summary(m1_30)
# pairwise significance
#emmeans(m1_30, list(pairwise ~ name_compress), adjust = "tukey")$`pairwise differences of name` %>% kable()
summary(glht(m1_30, linfct=c("name_compressSeedsfromberry - name_compressSeedsfromscat == 0",
                          "name_compressSeedsfromscat - name_compressMixedscat == 0")))


m1_60 <- glmer(prop.germ ~ name_compress + (1|id/jiffy), family = "binomial", weights = n_seeds, data = df%>%filter(days==60))
summary(m1_60)
# pairwise significance
#emmeans(m1_60, list(pairwise ~ name_compress), adjust = "tukey")$`pairwise differences of name` %>% kable()
summary(glht(m1_60, linfct=c("name_compressSeedsfromberry - name_compressSeedsfromscat == 0",
                          "name_compressSeedsfromscat - name_compressMixedscat == 0")))

```

```{r double check, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
### Double check that model is same regardless of specification (all good)
## this time don't weight, just include raw data
y <- cbind(df$total_seedlings, df$n_seeds - df$total_seedlings)
model <- glmer(y ~ name_compress + days.factor + (1|id/jiffy), family = "binomial", data = df)

# plot odds ratios
summary(model)

sjPlot::plot_model(model,
  show.values = TRUE, show.p = TRUE,
  title = "Effect of Treatment on Germination, w/ Random Effect for Sample"
)


# pairwise significance
summary(glht(model, linfct=c("name_compressSeedsfromberry - name_compressSeedsfromscat == 0",
                          "name_compressSeedsfromscat - name_compressMixedscat == 0"))) ##identical


```

# Germination Conclusion
Results suggest that A was much worse than all treatments (p<0.001), C was better than D (p<0.005), and B was better than D (p=0.05). We did not detect a significant difference between B and C (p=0.24). Qualitatively B germinated more than C, and the effect size was actually larger than for C to D, but statistically we didn't have the sample size/precision to resolve the difference between B and C.


#Task 2: Analyze dispersal potential using collar data and gut retention times

## load and clean telemetry data
```{r Load and clean telem data, results='hide', message=FALSE, warning=FALSE}
ev.grizz <- read.csv(here::here("data","EVcollar_Relocs_raw.csv"))

ev.grizz<- ev.grizz%>%
  mutate(datetime=DateTime%>%ymd_hms(tz="MST"),
         sp="grizzlybear",
         study="EV_Grizz_Lamb",
         fixtype="3D",
         sensortype="gps",
         Name=as.character(Name),
         id_yr=paste(Name,year(datetime),sep="_"),
         season=case_when(month(datetime)<8~"early",TRUE~"late"),
         id_yr_ssn=paste(Name,year(datetime),season,sep="_"))%>%
  filter(month(datetime)%in%4:11, DOP<=10)%>%
  dplyr::select(sp,
                study,
                id=Name,
                id_yr,
                id_yr_ssn,
                datetime,
                lat=Latitude,
                long=Longitude,
                dop=DOP,
                fixtype,
                sensortype)%>%
  tibble()%>%
  mutate(loc.id <- 1:nrow(.))%>%
  st_as_sf(coords=c("long","lat"),
           crs=4326)%>%
  st_transform("+proj=aea +lat_0=45 +lon_0=-126 +lat_1=50 +lat_2=58.5 +x_0=1000000 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")

##keep recent bears
ev.grizz <- ev.grizz%>%filter(year(datetime)>2015)

##keep only animal-years with >25 locations
keep <- ev.grizz%>%
  group_by(id_yr)%>%
  summarize(n=n())%>%
  filter(n>25)%>%
  pull(id_yr)

ev.grizz <- ev.grizz%>%filter(id_yr%in%keep)
```

## extract huckleberry data to points
```{r Huck to points, results='hide', message=FALSE, warning=FALSE}
##huck
huck <- terra::rast(here::here("data/VACCMEM_kcal.tif"))
#plot(elev)

##convert to terra raster 
pts.terra <- ev.grizz%>%terra::vect()

# extract
ev.grizz <-
  ev.grizz %>%
  mutate(huck = terra::extract(huck, pts.terra)[,2],
         month=month(datetime))


##summarise
ev.grizz%>%
  tibble%>%
  group_by(month)%>%
  summarise(mean=mean(huck))

#hist(ev.grizz$huck)

##select huck bears
huck.bears <-
  ev.grizz%>%
  tibble%>%
  filter(month==8)%>%
  group_by(id_yr)%>%
  summarise(mean=mean(huck))

#hist(huck.bears$mean)

huck.bears.keep<-huck.bears%>%
  filter(mean>800)

ev.grizz.huck<- ev.grizz%>%
  filter(month%in%c(8:9) & id_yr%in%huck.bears.keep$id_yr) ##select Aug and Sept
```


## analyze movements
```{r mvmt, results='hide', message=FALSE, warning=FALSE}
##get step length and fix rate
dist <- ev.grizz.huck%>%
  group_by(id_yr)%>%
  arrange(datetime)%>%
  mutate(time=as.numeric(lead(datetime)-datetime,"hours"),
         distance=st_distance(lead(geometry), geometry, 
                              by_element = TRUE)%>%as.numeric())%>%
  filter(time<20)

#hist(dist$time)

dist <- dist%>%
  mutate(time6=case_when(as.numeric(lead(datetime,n=3)-datetime,"hours")%>%round(0)%in% 5:7~as.numeric(lead(datetime,n=3)-datetime,"hours"),
                          as.numeric(lead(datetime,n=1)-datetime,"hours")%>%round(0)%in% 5:7~as.numeric(lead(datetime,n=1)-datetime,"hours")),
         distance6=case_when(as.numeric(lead(datetime,n=3)-datetime,"hours")%>%round(0)%in% 5:7~st_distance(lead(geometry,n=3), geometry, 
                                                                                                               by_element = TRUE)%>%as.numeric(),
                              as.numeric(lead(datetime,n=1)-datetime,"hours")%>%round(0)%in% 5:7~st_distance(lead(geometry,n=1), geometry, 
                                                                                                               by_element = TRUE)%>%as.numeric()),
         
         time15=case_when(as.numeric(lead(datetime,n=7)-datetime,"hours")%>%round(0)%in% 12:18~as.numeric(lead(datetime,n=7)-datetime,"hours"),
                          as.numeric(lead(datetime,n=2)-datetime,"hours")%>%round(0)%in% 12:18~as.numeric(lead(datetime,n=2)-datetime,"hours")),
         distance15=case_when(as.numeric(lead(datetime,n=7)-datetime,"hours")%>%round(0)%in% 12:18~st_distance(lead(geometry,n=7), geometry, 
                                                                                                               by_element = TRUE)%>%as.numeric(),
                              as.numeric(lead(datetime,n=2)-datetime,"hours")%>%round(0)%in% 12:18~st_distance(lead(geometry,n=2), geometry, 
                                                                                                               by_element = TRUE)%>%as.numeric())
  )


hist(dist$huck)
hist(dist$distance6)
hist(dist$distance15)


##get average step length
dist_stepped <- dist%>%
  tibble%>%
  filter(huck>1000)%>%
  group_by(month,id_yr)%>%
  mutate(speed=distance/time)%>%
  summarise_at(c("speed","time","distance"), mean, na.rm = TRUE)

# dist_stepped%>%
#   group_by(month)%>%
#   summarise_at(c("speed","time","distance"), mean, na.rm = TRUE)


dist_stepped6 <- dist%>%
  tibble%>%
  filter(huck>1000)%>%
  group_by(month,id_yr)%>%
  mutate(speed6=distance6/time6)%>%
  add_count()%>%
  summarise_at(c("time6","distance6","n"), c(mean=mean,max=max), na.rm = TRUE)%>%
  select(-n_max)%>%
  rename(n=n_mean)




dist_stepped6%>%
  select(-n, -contains("time"))%>%
  group_by(month)%>%
  summarise_at(c("distance6_mean","distance6_max"), c(mean=mean,sd=sd), na.rm = TRUE)


dist_stepped6.final <- dist_stepped6%>%
  select(-n)%>%
  ungroup%>%
  summarise_at(c("distance6_mean","distance6_max","time6_mean", "time6_max"), c(mean=mean,sd=sd), na.rm = TRUE)



dist_stepped15 <- dist%>%
  tibble%>%
  filter(huck>1000)%>%
  group_by(month,id_yr)%>%
  mutate(speed15=distance15/time15)%>%
  add_count()%>%
  summarise_at(c("time15","distance15","n"), c(mean=mean,max=max), na.rm = TRUE)%>%
  select(-n_max)%>%
  rename(n=n_mean)
  

dist_stepped15%>%
  select(-n, -contains("time"))%>%
  group_by(month)%>%
  summarise_at(c("distance15_mean","distance15_max"), c(mean=mean,sd=sd), na.rm = TRUE)


dist_stepped15%>%
  select(-n, -contains("time"))%>%
  ungroup%>%
  summarise_at(c("distance15_mean","distance15_max"), c(mean=mean,sd=sd), na.rm = TRUE)
```





## bootstrap
```{r bootstrap, results='hide', message=FALSE, warning=FALSE}
##bootstrap
boot.n <- 1000

##6 hour
dist_stepped6.dat <- dist%>%
  tibble%>%
  filter(huck>1000)%>%
  group_by(month,id_yr)%>%
  mutate(speed6=distance6/time6)%>%
  drop_na(speed6)%>%
  ungroup

dist_stepped6.boot <-tibble()
for(i in 1:boot.n){
a <- dist_stepped6.dat%>%
  sample_frac(1, replace=TRUE)%>%
  group_by(id_yr)%>%
  summarise(speed_yr=mean(speed6),
            speed_yr_max=max(speed6))%>%
  ungroup%>%
  summarise(speed=mean(speed_yr),
            speed_max=mean(speed_yr_max))%>%
  mutate(dist6=speed*6.25,
         dist6_max=speed_max*6.25)

dist_stepped6.boot <- rbind(dist_stepped6.boot,a)
}

dist_stepped6.boot%>%
  summarise_at(c("speed","dist6","dist6_max"), c(mean=mean,sd=sd), na.rm = TRUE)


##15 hour

dist_stepped15.dat <- dist%>%
  tibble%>%
  filter(huck>1000)%>%
  group_by(month,id_yr)%>%
  mutate(speed15=distance15/time15)%>%
  drop_na(speed15)%>%
  ungroup

dist_stepped15.boot <-tibble()
for(i in 1:boot.n){
  a <- dist_stepped15.dat%>%
    sample_frac(1, replace=TRUE)%>%
    group_by(id_yr)%>%
    summarise(speed_yr=mean(speed15),
              speed_yr_max=max(speed15))%>%
    ungroup%>%
    summarise(speed=mean(speed_yr),
              speed_max=mean(speed_yr_max))%>%
    mutate(dist15=speed*15.6,
           dist15_max=speed_max*15.6)
  
  dist_stepped15.boot <- rbind(dist_stepped15.boot,a)
}

dist_stepped15.boot%>%
  summarise_at(c("speed","dist15","dist15_max"), c(mean=mean,sd=sd), na.rm = TRUE)
```

